---
title: "MATH3091: Statistical Modelling II"
subtitle: "Problem Sheet 5"
format: pdf
header-includes:
  - \usepackage{xcolor}
  - \definecolor{solutionblue}{RGB}{0, 0, 255}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


1. The time to failure ($Y$) of a certain type of electrical component is thought to follow an exponential distribution, with probability density of the form
$$
f_Y(y;\lambda) = \lambda \exp\left(-\lambda y\right), \quad
y>0;\quad\lambda>0\,.
$$
It is believed that the failure rate of a component $\lambda$ is related to its electrical resistance ($x$) by the relationship
$$
\lambda = \beta_1+\beta_2 x\,.
$$
Suppose that $y_1,\cdots ,y_n$  are observations of the times to failure, $Y_1,\cdots,Y_n$ for $n$ such components with corresponding
resistances $x_1,\cdots,x_n$.

   a. Write down the likelihood in terms of $\beta_1$ and $\beta_2$ and hence derive a pair of simultaneous equations, the solutions of which are the maximum likelihood estimates.

   b. Calculate the observed and expected information matrices. Are the Newton-Raphson and the Fisher scoring methods identical
for this problem? Justify your answer.

2. Suppose $Y_i \sim \text{Geometric}(p_i)$, the geometric distribution as studied in Question 2 of Problem Sheet 4. We want to model how $p_i$ depends on explanatory variables $\boldsymbol{x}_i$.

   a. Assuming a GLM with canonical link function, write down a formula for $p_i$ in terms of $\boldsymbol{x}_i$. Is this a sensible model?

   a. Suppose instead that 
   $$
   \text{logit} (p_i) = \log \frac{p_i}{1 - p_i} =  \mathbf{x}_i^{\top} \boldsymbol{\beta}\,.
   $$
   Show that this is a GLM with a non-canonical link function, and write down the link function corresponding to this model.
   
   b. Derive an expression for the scaled deviance for this model, writing $\hat \mu_i$ for the estimate of $\mu_i = \mathbb{E}(Y_i)$ under the model from part (b). Write an expression for $\hat \mu_i$ in terms of $\hat{\boldsymbol{\beta}}$, the MLE of $\boldsymbol{\beta}$.

3. We return to the `beetle` data studied in Computer Lab 5, with observations on $n = 8$ groups of beetles. There we considered the model:

```{r, include = FALSE}
beetle <- read.csv("beetle.csv")
beetle$prop_killed <- beetle$killed / beetle$exposed
```

```{r}
beetle_glm <- glm(prop_killed ~ dose, data = beetle, family = binomial,
                  weights = exposed)
```

We could have also considered a model with quadratic dependence on `dose`
```{r}
beetle_glm_quad <- glm(prop_killed ~ dose + I(dose^2), data = beetle, 
                       family = binomial, weights = exposed)
```

   a. Write down mathematical expressions for the two models. Show that `beetle_glm` is nested with `beetle_glm_quad`, and write down the null hypothesis $H_0$ and the alternative hypothesis $H_1$ you would use for comparing the models.
   
   b. Consider the following output of a `summary()` call. What is the scaled deviance for `beetle_glm`? 
```{r}
summary(beetle_glm)
```
    
   c. The scaled deviance for `beetle_glm_quad` is $3.1949$. Calculate the log likelihood ratio test statistic $L_{01}$ for testing $H_0$ against $H_1$. Under $H_0$, what is the distribution of this statistic? Hence conduct a hypothesis test of $H_0$ against $H_1$, and make a conclusion about which model you prefer. 
