---
title: "MATH3091: Computer Lab 08"
format: 
  html:
    css: styles.css
---

```{r setup, echo = FALSE, include = FALSE}
#| echo: TRUE
#| output: TRUE
#| message: FALSE

knitr::opts_chunk$set(include = TRUE, fig.width = 6, fig.height = 4, fig.align = "center")
```

## Analysis of `accident.csv` data

Load the data in `accident.csv` into a variable `accident`.
```{r}
accident <- read.csv("../datasets/accident.csv") # Change the path, if needed
head(accident)
```

The data in `accident.csv` concerns the number of road accidents (`number`) and the volume of traffic (`volume`), on each of two roads in Cambridge (`road`), at various times of day (`time`, taking values `morning`, `midday` or `afternoon`).

We would like to answer questions like:

1. Is Mill Road more dangerous than Trumpington Road?
1. How does time of day affect the rate of road accidents? 

Let $Y_{ij}$ denote the number of accidents and $\mu_{ij}$ denote the expected number of accidents for the $i$th road at the $j$th time. We assume

$$
Y_{ij} \sim \text{Poisson}(\mu_{ij}), \ i=1, 2 \text{ for Road},
j=1, 2, 3 \text{ for time}.
$$


We might reasonably expect the number of accidents to depend on traffic `volume`, $v_{ij}$.  It is better to work with log volume rather than volume itself as the volumes are very high.

Use the Poisson GLM with canonical log link.

$$
\log (\mu_{ij}) = \mu + \alpha_i + \beta_j + \gamma \log v_{ij}.
$$
This is equivalent to assuming:

$$
\mu_{ij} = \text{constant} \times \text{$i$-th road effect}
  \times \text{$j$0th time effect}  \times \text{volume}^{\gamma}
$$

By default, `R` will set $\alpha_1=0$ and $\beta_1=0$. Thus $\alpha_2=0$ if the roads are equally risky, $\beta_2$ represents difference between time 2 and 1,  and $\beta_3$ represents difference between time 3 and 1.

We can fit this model in `R` as follows.
```{r}
# Fit a Poisson log-linear model to the data:
acc_glm <- glm(number ~ road + time + log(volume), family = poisson, data = accident)
summary(acc_glm)
```

- [What do you conclude about how the accident rate depends on the road, the time of day and the volume of traffic?]{.attention}

```{r, include = FALSE}
# The output seems to say Mill road is more dangerous than Trumpington road. 
# The mornings and afternoons are about as dangerous as each other and each is quite a lot more dangerous than the middle of the day. 
# The accident rate has a strong dependence on the traffic volume.
```

- [Does the scaled deviance indicate an problem with the fit of the model?]{.attention}

```{r, include = FALSE}
# From the summary output, the scaled deviance is 1.88. 
# `acc_glm` has 5 free parameters, and there are `n = 6` data points, so if the model is correct, we expect the scaled deviance to have chi^2_1 distribution.
# The 95% point of a chi^2_1 is
qchisq(0.95, df = 1)
# i.e., 3.84. 
# Since the scaled deviance is less than this, there is no evidence of poor model fit from this measure.
```

- [Conduct hypothesis tests to check whether you could drop any terms from the model.]{.attention}

```{r}
# Consider model without `volume`.
acc_glm_rt <- glm(number ~ road + time, family  = poisson, data = accident)
anova(acc_glm_rt, acc_glm, test="LRT")
# We prefer the model with `volume`.

# Consider model without `time`.
acc_glm_rv <- glm(number ~ road + log(volume), family = poisson, data = accident)
anova(acc_glm_rv, acc_glm, test="LRT")
# It is borderline whether or not to include `time`.

# Consider model without `road`.
acc_glm_tv <- glm(number ~ time + log(volume), family = poisson, data = accident)
anova(acc_glm_tv, acc_glm, test="LRT")
# We prefer the model with `road`.

```

## Analysis of `hodgkins.csv` data

The data in `hodgkins.csv` concerns 538 patients with Hodgkin's disease, who have been cross-classified according to two factors, `type` the histological type of their disease (4 levels) and `rtreat`, their response to treatment (3 levels).

Load the data in `hodgkins.csv` into a variable `hodgkins`.
```{r}
# Analysis of `hodgkins.csv` data
hodgkins <- read.csv("../datasets/hodgkins.csv") # Change the path, if needed
head(hodgkins)
```

We could display this data in a contingency table.
```{r}
# Display the data in contingency table form:
xtabs(count ~ type + rtreat, data = hodgkins)
```

Cross-classification data are extremely common, and can be modelled very effectively using generalised linear models.
The observations of the response variable are taken to be the counts (in this case the 12 patient totals) and a generalised
linear model is used to determine how the expected counts depend on any explanatory variables (in this case, the factors `type` and `rtreat`). The dataset as it is presented in the original `hodgkins` data frame is already in the correct format to fit
such a generalised linear model.

Counts are non-negative integers, so one approach is to treat them as observations of Poisson random variables. The canonical link function is then the $\log$ function, and Poisson generalised linear models with the log link are called **log-linear models**.

A possible log-linear model for this data set is
$$
Y_i\sim\text{Poisson}(\mu_i),\quad
  \log\mu_i=\alpha+\beta_T(t_i)+\beta_R(r_i)\qquad i=1,\cdots ,12\,,
$$
where $Y_i$ is the $i$-th count and $t_i$ and $r_i$ are the corresponding levels of `type` and `rtreat`.

We can fit this model in `R`.
```{r}
# Fit a log-linear model with main effects only.
hod_main <- glm(count ~ type + rtreat, family = poisson, data = hodgkins)
```

This model only contains the "main effects" `type` and `rtreat`. Alternatively, we could fit a model with an interaction between these two main effects, which is the saturated model.
```{r}
# Fit a log-linear model with interaction between main effects.
hod_int <- glm(count ~ type * rtreat, family = poisson, data = hodgkins)
```

We can find the estimated cell counts under each of the models.
```{r}
# Find estimated cell counts under the two models.
fitted(hod_main)
fitted(hod_int)
```

[What do you notice about the estimated cell counts under the saturated model `hod_int`, compared with the observed counts?]{.attention}
```{r, include = FALSE}
# For the saturated model, the estimated cell counts are equal to the observed cell counts.
```

[Which model do you prefer?]{.attention}
```{r}
# Compare the two models
anova(hod_main, hod_int, test = "LRT")
```

```{r, include = FALSE}
# There is strong evidence to prefer `hod_int`, and we conclude that response to treatment is dependent on the type of the disease.
```

By comparing these two models, we are determining whether `type` and `rtreat` are independent, or whether there is significant evidence of association. If we need the interaction term, then this provides evidence that the cross-classifying factors are not independent of one another.
